The algebraic hierarchy in type theory using type classes

### 2010-06-23 jar.tex
Reintegrate small note about ssreflect.
Started work on quotation section.

% todo:
  % what to say about terms referring on proofs?
  % cite that other type class based development in the context of the alternative bundling schemes.
  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
  % say this somewhere: \footnote{For an interesting (if fairly restrictive) alternative design in which unicity of term representations \emph{is} required for all objects, see the algebraic hierarchy in the ssreflect\cite{ssreflect} standard library.}

@@

@@

% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for those that means only the operational classes, which we certainly don't think of as being part of the hierarchy.

@@

Following our idiom, we introduce operational type classes for the \emph{components} of a category:

\begin{lstlisting}
Class Arrows (O: Type): Type := Arrow: O → O → Type.
Class CatId O `{Arrows O} := cat_id: `(x ⟶ x).
Class CatComp O `{Arrows O} :=
  comp: Π {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:

\begin{lstlisting}
Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
    `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem in more detail:
\begin{theorem}[First homomorphism theorem]
If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
\end{theorem}

@@

% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), functors (including automatically generated forgetful functors), natural transformations, adjunctions, initial models of equational theories constructed from term algebras, transference of proofs between isomorphic models of equational theories, subalgebras, congruences, quotients, products, and the first homomorphism theorem~\cite{meinke1993universal}.

@@

%The product of two $\tau$-algebras is their categorical product.
% 
% \begin{comment}
% Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
% An interesting fact to mention is the use of the following heterogeneous equality between
% elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
% \begin{lstlisting}
%   Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
%     match o with
%     | constant x => fun a b => `a == b
%     | function x y => fun a b => forall u, heq (a u) (b (`u))
%     end.
% \end{lstlisting}
% \end{comment}
% 
% We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
% of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
% Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
% %varieties/monoid.v
% \begin{lstlisting}
% Variable o: variety.Object theory.
% Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
% Global Instance: MonoidUnit (o tt) := algebra_op theory one.
% Global Instance from_object: Monoid (o tt).
% \end{lstlisting}
% 
% In the last line, instance resolution `automatically' finds the operation and the unit specified in
% the lines before.
% 
% This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic). However, it opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins. We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.
% 
%

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function.\footnote{Gonthier provides similar functionality using canonical structures.} This feature is absent from Coq. Fortunately, type classes provide similar functionality as we will now show.

We define a term language for monoids
\begin{lstlisting}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{lstlisting}

The expression type is parametrized over the set of variable indices. Hence, we diverge from~\cite{Hints}, 
which uses \lstinline|nat| for variables thereby introducing bound problems and dummy variables.

@@

Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
  match e with
  | Zero => 0
  | Mult a b => eval vs a * eval vs b
  | Var v => vs v
  end.
\end{lstlisting}

@@

We have used this technique to implement a tactic to normalize monoid expressions.
%It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
%the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
%which moreover works on setoid equalities.

>>>>

% todo:
  % what to say about terms referring on proofs?
  % cite that other type class based development in the context of the alternative bundling schemes.
  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)

@@

For an interesting (if fairly restrictive) alternative approach based on the idea of requiring unicity of representation for all objects so that Leibniz equality is always applicable, see the algebraic hierarchy part of the ssreflect\cite{ssreflect} libraries.

@@

% note: we don't mention "super classes" because by Matthieu's terminology only parameters-of-class-type are "super classes", and for us that means only the operational classes, which we certainly don't think of as being part of the hierarchy. Matthieu calls fields of class type declared with ":>" substructures.

@@

Following our idiom, we introduce operational type classes for the \emph{components} of a category:
\begin{lstlisting}
Class Arrows (O: Type): Type := Arrow: O → O → Type.
Class CatId O `{Arrows O} := cat_id: `(x ⟶ x).
Class CatComp O `{Arrows O} :=
  comp: Π {x y z}, (y ⟶ z) → (x ⟶ y) → (x ⟶ z).

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
    `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem~\cite{meinke1993universal} in more detail:
\begin{theorem}[First homomorphism theorem]
If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
\end{theorem}

@@

% Eelis: TODO: process: At the time of writing, our development includes a fully integrated formalization of a nontrivial portion of category theory and multisorted universal algebra, including various categories (e.g.\ the category \lstinline|Cat| of categories, and finitary algebraic categories  defined by a theory which we instantiate to obtain the categories of monoids, semirings, and rings), products

@@

%The product of two $\tau$-algebras is their categorical product.
% 
% \begin{comment}
% Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
% An interesting fact to mention is the use of the following heterogeneous equality between
% elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
% \begin{lstlisting}
%   Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
%     match o with
%     | constant x => fun a b => `a == b
%     | function x y => fun a b => forall u, heq (a u) (b (`u))
%     end.
% \end{lstlisting}
% \end{comment}
% 
% We connect the abstract theory to the concrete theory. Concretely, let $\tau_m$ be the theory
% of monoids. Given a (concrete) monoid, we can construct the corresponding $\tau_m$-algebra. 
% Conversely, given a $\tau_m$-algebra, we construct an instance of the \lstinline|Monoid| type class.
% %varieties/monoid.v
% \begin{lstlisting}
% Variable o: variety.Object theory.
% Global Instance: SemiGroupOp (o tt) := algebra_op theory mult.
% Global Instance: MonoidUnit (o tt) := algebra_op theory one.
% Global Instance from_object: Monoid (o tt).
% \end{lstlisting}
% 
% In the last line, instance resolution `automatically' finds the operation and the unit specified in
% the lines before.
% 
% This interplay between concrete algebraic structure and their expressions on the one hand, and models of equational theories on the other is occasionally a source of tension (in that translation in either direction is not yet fully automatic). However, it opens the door to the possibility of fully internalized implementations of generic tactics for algebraic manipulation, no longer requiring plugins. We come back to this when we describe automatic quotation of concrete expressions into universal algebra expressions; see Section~\ref{quote}.

\section{Quoting with type classes}\label{quote}

A common need when interfacing generic theory and utilities developed for algebraic structures (such as normalization procedures) with concrete instances of these structures is to take a concrete expression or statement in a model of a particular algebraic structure, and translate it to a symbolic expression or statement in the language of the algebra's signature, so that its structure can be inspected.

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''. In~\cite{Hints}, TODO et. al. show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality using ingenious use of canonical structures.} This feature is absent from Coq, but fortunately type classes let us do very much the same thing, as we will now show.

For ease of presentation we show only a proof of concept for a very concrete language. We are currently working to integrate this technique with our existing universal algebra infrastructure, in particular its data types for algebraic terms.%  Preliminary results in this area (such as a simple monoid normalization tactic) suggest yada.

For the present example, we define a term language for monoids
\begin{lstlisting}
Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}

The expression type is parametrized over the set of variable indices. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.

@@

Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
  match e with
  | One => 1
  | Mult a b => eval vs a * eval vs b
  | Var v => vs v
  end.
\end{lstlisting}

@@

We have used this technique to implement a tactic to normalize monoid expressions, and plan to extend yada yada.
%It would be interesting to use this technique to implement the \lstinline|congruence| tactic, which implements
%the congruence closure algorithm~\cite{corbineau2007deciding}. One would naturally obtain a tactic
%which moreover works on setoid equalities.
